Done in QGIS and PowerPoint to keep it simple

```{r}
library(terra)
library(sf)
library(dplyr)
library(tmap)
library(here)

fig_widths <- c(min = 1.18, single = 3.48, mid = 5.51, two = 7.16)

# run to save projected roads for use in QGIS
p_rds_mst <- readRDS(here::here("analysis/data/derived_data/bench_results/result_mst_regular_1e-05_10_revelstoke.rds"))

write_sf(p_rds_mst, here::here("analysis/data/derived_data/bench_results/mst_regular_1e-05_10_revelstoke.gpkg"))
```


In QGIS open input_cutblocks and input_roads for TSA 27 as well as combined_revelstoke_roads and projected roads for all cutblocks.  Get Google satellite imagery from XYZ tile service. 

Assign colors to as follows
Existing roads: #2c7bb6
Projected roads: #d7191c
Observed roads: #0571b0 "#fdae61"
Cutblocks: #92c5de
               
Then capture images in the map views desired by clicking Project > Import/Export > Export map as image,
setting the dpi to 300 and saving as png. 

To make the raster map we need to manipulate some of the rasters.

```{r}
data_path_raw <- "analysis/data/raw_data/"
data_path_drvd <- "analysis/data/derived_data/TSA27/dem"

dist_obs <- rast(here::here(file.path(data_path_drvd, "distanceToRoadobserved_NA_NA.tif")))

dist_proj <- rast(here::here(file.path(data_path_drvd, "distanceToRoadregular_1e-05_mst.tif")))

dist_dif <- dist_obs - dist_proj

# need to clip/mask everything to the tsa boundary to avoid problems outside tsa

tsaBoundary <- read_sf(here(data_path_raw, "tsa27_boundaries.gpkg")) %>% st_union()

dist_dif <- mask(dist_dif, vect(tsaBoundary))

e_rds <- st_read(here::here(file.path(data_path_drvd, "../input_roads.gpkg")))
p_rds <- readRDS(here::here("analysis/data/derived_data/bench_results/result_mst_regular_1e-05_10_revelstoke.rds"))
o_rds <- st_read(here::here(file.path(data_path_drvd, "../../combined_revelstoke_roads.gpkg")))

rds <- lst(e_rds, p_rds, o_rds) %>% 
  purrr::map(st_intersection, tsaBoundary)

mp <- tm_shape(dist_dif)+
  tm_raster(col.scale = tm_scale_continuous(
    labels = c("<= -5000", "     -2500", "      2500", ">=  5000"),
    values = "brewer.prgn",
    ticks = c(-5000, -2500, 2500, 5000), 
    midpoint = 0
  ), 
  col.legend = tm_legend(title = "Observed - projected\ndistance to road", 
                         title.padding = c(0.01, 0,0,0),
                         title.size = 0.75,
                         height = 8, frame = FALSE))+
  tm_shape(rds$o_rds)+
  tm_lines(col = "black", lty = "12")+
  tm_shape(rds$p_rds)+
  tm_lines(col = "#fdae61")+
  tm_shape(rds$e_rds)+
  tm_lines(col = "black")+
  tm_layout(legend.position = c("LEFT", "BOTTOM"), legend.just = "left", frame = FALSE)+
  tm_add_legend(type = "lines", col = c("black", "#fdae61", "black"),
                lty = c("solid","solid","dotted"),
                lwd = 2,
                labels = c("Existing roads", "Projected roads", "Observed roads"))+
  tm_add_legend(type = "polygons", fill = "white", col = "#2c7bb6", labels = "Cutblocks",
                border.lwd = 2)

tmap_save(mp,
          here("analysis/figures/spatial_projections.png"),
          dpi = 300, height = 7, width = fig_widths["one"])


# p_rds_ilcp <- readRDS(here::here("analysis/data/derived_data/bench_results/result_ilcp_regular_1e-05_10_revelstoke_real.rds"))
# 
# write_sf(p_rds_ilcp, here::here("analysis/data/derived_data/bench_results/ilcp_regular_1e-05_10_revelstoke_real.gpkg"))
#  p_rds_mst <- readRDS(here::here("analysis/data/derived_data/bench_results/result_mst_regular_1e-05_10_revelstoke_real.rds"))
# 
# write_sf(p_rds_mst, here::here("analysis/data/derived_data/bench_results/mst_regular_1e-05_10_revelstoke_real.gpkg"))
```

Then arrange the images in powerpoint, modifying the contrast slightly to make the colours show on the satellite imagery. Save to tif by highlighting all elements > right click > save as picture > choose tiff 
